Quantum Hoare Type Theory

### 2020-10-26 ms-paper/ms-paper.tex
Convert booleans to bits

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \type{1} \mid \type{Bool} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Assertions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid M \cdot P \mid X \in_q S \mid X =_q \psi \mid \kw{uniform}(X) \mid \kw{separable}(X) \mid \kw{classical}(X) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{true} \mid \kw{false} $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \kw{init} N \mid \kw{apply} U \kw{to} N \mid \kw{measure} N \mid \kw{if} y \kw{then} E \kw{else} F$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
		\textit{Assertion context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

\paragraph{Types} We include primitive types for unit, booleans, qubits and unitary operations, and a type \type{Pure} for representing pure quantum state vectors (complex valued vectors in Hilbert space); these correspond to similar types in the QIO work. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.

@@

\paragraph{Quantum Commands}
These are the quantum-specific effectful commands inspired by QIO for initialization, unitary application and measurement of qubits.
\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically 0 or 1 if N is a boolean).
\texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
\texttt{measure} $x$ measures the variable $x$ in the computational basis.
\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Bool}
		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(N)) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
\end{mathpar}
where x is fresh and \texttt{state} is a function that translates a classical representation of quantum state (such as boolean here) to the equivalent quantum state vector.

Initialization can only be performed for a term $N$ that can be type checked as a \type{Bool}. Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ points to the representation of a new qubit state; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Qbit}
		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
		\\ \Delta, x:\type{Bool}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bool}.Q)}
\end{mathpar}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bool {\top /\ Id(r, false)}
    = do q <= init false;
         return measure q
\end{lstlisting}
\end{minipage}

In this trivial program where we initialize a new qubit with \texttt{false} and then immediately measure it, we assert that the result is equal to \texttt{false}. The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ assertion.

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bool {\top /\ Id(r, false)}
-- P0: \top
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         return measure q
-- P2: P1 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bool {\top}
    = do q <= init false;
         apply H to q;
         return measure q
\end{lstlisting}
\end{minipage}

In the case of quantum coin toss, the result can be in either of its two boolean values \texttt{false} or \texttt{true}, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bool {\top}
-- P0: \top
    = do q <= init false;
-- P1: P0 \o (q |-> |0\>)
         apply H to q;
-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
         return measure q
-- P3: P2 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

\subsubsection{Testing Bell Pair}
\leavevmode
\todo{TODO: Should the type be \type{Bool} $\otimes$ \type{Bool}?}
\todo{TODO: syntax for applying multi-qubit gate}

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}  -- a,b \in {true, false}
         = do qa <= init false;
              apply H to qa;
              qb <= init false;
              apply CNOT to [qa, qb];
              return (measure qa, measure qb)
\end{lstlisting}
\end{minipage}

Here, we are asserting that the returned booleans $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program.

@@

\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool, Bool) {\top /\ Id(a, b)}
              -- P0: \top
         = do qa <= init false;
              -- P1: P0 \o (qa |-> |0\>)
              apply H to qa;
              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
              qb <= init false;
              -- P3: P2 \o (qb |-> |0\>)
              apply CNOT to [qa, qb];
              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
              return (measure qa, measure qb)
              -- P5: P4 \o ((qa |-> -) -o \top) \o ((qb |-> -) -o \top)
\end{lstlisting}

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=init]
				{\inferrule*[right=measure]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bool} \vdash r \chk \type{Bool}}
						{q : \type{Qbit}, r : \type{Bool}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bool} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bool} . (\exists r : \type{Bool}.R)}
					\\ \cdot \vdash \kw{false} \chk \type{Bool}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{init false}; E_{23} \syn r : \type{Bool} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bool}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{false})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bool} . (\top \multimap \top \wedge Id(r, \kw{false}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bool} \{\top \wedge \kw{Id}(r, \kw{false})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {\top} r : Qbit {Id(r, |+\>)}
      = do q <= init false;
           apply H to q;
           return q

qminus : {\top} r : Qbit {Id(r, |-\>)}
       = do q <= init true;
            apply H to q;
            return q
\end{lstlisting}

\subsection{Creating entanglement}
\begin{lstlisting}[language=QHaskell]
share : \Pi a : Qbit.
        {a \in {|+\>, |-\>}}   -- a \in {b, c} is short for Id(a,b) \/ Id(a,c)
           b : Qbit
        {Id(a, b) /\ a \in {|0\>, |1\>}}
      = \a.do b <= init false;
              apply CNOT to [a, b];
              return b
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bool o* Bool) {\top /\ Id(a, b)}
         = do (qa, qb) <- bell;
              return (measure qa, measure qb)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi a : Qbit. \Pi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bool o* Bool)
        {\top}
      = \a.\e.do apply CNOT to [a, e];
                 apply H to a;
                 return (measure a, measure e)
\end{lstlisting}

\subsection{Bob's circuit}
\begin{lstlisting}[language=QHaskell]
bob : \Pi m1 m2 : Bool. \Pi e : Qbit
      {entangled(e)}
         r : Qbit
      {(Id(r, -)}
      = \m1.\m2.\e.do if m1 then (apply Z to e) else ();
                      if m2 then (apply X to e) else ();
                      return e
\end{lstlisting}

>>>>

\begin{figure}[ht]
	\begin{tabular}{lrcl}
		\textit{Types} & $A, B, C$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \type{1} \mid \type{Bit} \mid \type{Qbit} \mid \type{U} \mid A \otimes B \mid \Pi x{ : }A.B \mid \Delta. \Psi .\{P\} x{ : }A \{Q\}$
		\end{minipage}\\ \\
		\textit{Assertions} & $P, Q, R$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \top \mid \bot \mid P \wedge Q \mid P \vee Q \mid M \cdot P \mid X \in_q S \mid X =_q \psi \mid \kw{uniform}(X) \mid \kw{separable}(X) \mid \kw{classical}(X) $
		\end{minipage}\\ \\
		% \textit{Q. Heaps} & $H, G$ & ::= & $ h \mid \kw{empty} \mid \kw{upd}(H, M, N) $\\ \\
		\textit{Elim terms} & $K, L$ & ::= & $ x \mid K N \mid N : A $\\
		\textit{Intro terms} & $N, O$ & ::= & $ K \mid () \mid \lambda x.N \mid \kw{do} E \mid \kw{1} \mid \kw{0} $\\ \\
		\textit{Q. Commands} & $c$ & ::= &
		\begin{minipage}[t]{0.45\columnwidth}%
			$ \kw{init} N \mid \kw{apply} U \kw{to} N \mid \kw{measure} N \mid \kw{if} y \kw{then} E \kw{else} F$
		\end{minipage}\\
		\textit{Computations} & $E, F$ & ::= & $ \kw{return} N \mid x \leftarrow K; E \mid x \Leftarrow c; E \mid x =_A N; E $\\ \\
		\textit{Variable context} & $\Delta$ & ::= & $ \cdot \mid x : A, \Delta $ \\
		\textit{Q. Heap context} & $\Psi$ & ::= & $ \cdot \mid h, \Psi$ \\
		\textit{Assertion context} & $\Gamma$ & ::= & $ \cdot \mid P, \Gamma$ \\
	\end{tabular}
	\Description{Syntax of Quantum Hoare Type Theory}
	\caption{Syntax of Quantum Hoare Type Theory}
	\label{fig:syntax}
\end{figure}

\paragraph{Types} We include primitive types for unit, bits, qubits and unitary operations, and a type \type{Pure} for representing pure quantum state vectors (complex valued vectors in Hilbert space); these correspond to similar types in the QIO work. There are also type constructors for pairs, dependent functions and the Hoare type from HTT.

@@

\paragraph{Quantum Commands}
These are the quantum-specific effectful commands inspired by QIO for initialization, unitary application and measurement of qubits.
\texttt{init} $N$ initializes a given variable based on the value of the expression $N$ (typically 0 or 1 if N is a bit).
\texttt{apply} $U$ \texttt{to} $X$ applies unitary operation $U$ to variables $X$.
\texttt{measure} $x$ measures the variable $x$ in the computational basis.
\texttt{if y then E else F} executes computation $E$ if measuring variable $y$ returns 1 else executes $F$.

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Bit}
		\\ \Delta, x : \type{Qbit}; P \circ (x \mapsto \kw{state}(N)) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{init} N; E \syn y:B. (\exists x:\type{Qbit}.Q)}
\end{mathpar}
where x is fresh and \texttt{state} is a function that translates a classical representation of quantum state (such as boolean here) to the equivalent quantum state vector. \todo{TODO}

Initialization can only be performed for a term $N$ that can be type checked as a \type{Bit}. Then, we look at the rest of the computation, $E$, in a context that adds $x$ of type \type{Qbit} under a precondition that extends the previous precondition, $P$, with the strongest postcondition for initialization, that is, the newly bound variable $x$ points to the representation of a new qubit state; this should synthesize the strongest postcondition (with respect to the expanded precondition in the context), $Q$, for $E$.

@@

\begin{mathpar}
	\inferrule
	{\Delta \vdash N \chk \type{Qbit}
		\\ \Delta; \Psi; P \prove (N \hookrightarrow -)
		\\ \Delta, x:\type{Bit}; P \circ ((N \mapsto -) \multimap \top) \vdash E \syn y:B . Q}
	{\Delta; P \vdash x \Leftarrow \kw{measure} N; E \syn y:B. (\exists x:\type{Bit}.Q)}
\end{mathpar}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bit {\top /\ r=0}
    = do q <= init 0;
         return measure q
\end{lstlisting}
\end{minipage}

In this trivial program where we initialize a new qubit with 0 and then immediately measure it, we assert that the result is equal to 0. The complete specification also implicitly says that the existing quantum state is not affected in any way as both the pre and postcondition include the $\top$ assertion.

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
hqw : {\top} r : Bit {\top /\ r=0}
-- P0: \top
    = do q <= init 0;
-- P1: P0 \o (q |-> |0\>)
         return measure q
-- P2: P1 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bit {\top}
    = do q <= init 0;
         apply H to q;
         return measure q
\end{lstlisting}
\end{minipage}

In the case of quantum coin toss, the result can be in either 0 or 1, hence, we need not specify any special postcondition as type checking the result will be sufficient for correctness.

@@

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
rnd : {\top} r : Bit {\top}
-- P0: \top
    = do q <= init 0;
-- P1: P0 \o (q |-> |0\>)
         apply H to q;
-- P2: P1 \o ((q |-> |0\>) -o (q |-> |+\>))
         return measure q
-- P3: P2 \o ((q |-> -) -o \top)
\end{lstlisting}
\end{minipage}

\subsubsection{Testing Bell Pair}
\leavevmode
\todo{TODO: Should the type be \type{Bit} $\otimes$ \type{Bit}?}
\todo{TODO: syntax for applying multi-qubit gate}

\begin{minipage}{0.95\linewidth}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bit, Bit) {\top /\ Id(a, b)}  -- a,b \in {0, 1}
         = do qa <= init 0;
              apply H to qa;
              qb <= init 0;
              apply CNOT to [qa, qb];
              return (measure qa, measure qb)
\end{lstlisting}
\end{minipage}

Here, we are asserting that the returned bits $a$ and $b$ hold the same value, which is what we expect from the first Bell state prepared in this program. \todo{TODO}

@@

\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bit, Bit) {\top /\ Id(a, b)}
              -- P0: \top
         = do qa <= init 0;
              -- P1: P0 \o (qa |-> |0\>)
              apply H to qa;
              -- P2: P1 \o ((qa |-> |0\>) -o (qa |-> |+\>))
              qb <= init 0;
              -- P3: P2 \o (qb |-> |0\>)
              apply CNOT to [qa, qb];
              -- P4: P3 \o ((qa |-> |+\>, qb |-> |0\>) -o (qa, qb) |-> |\Phi+\>)
              return (measure qa, measure qb)
              -- P5: P4 \o ((qa |-> -) -o \top) \o ((qb |-> -) -o \top)
\end{lstlisting}

@@

\begin{turn}{90}
	\begin{minipage}{42pc}
	\begin{mathpar}
		\inferrule*[right=HIntro]
		{\inferrule*[right=consq]
			{\inferrule*[right=init]
				{\inferrule*[right=measure]
					{\inferrule*[right=return]
						{q : \type{Qbit}, r : \type{Bit} \vdash r \chk \type{Bit}}
						{q : \type{Qbit}, r : \type{Bit}; P_1 \circ ((q \mapsto -) \multimap \top) \vdash \kw{return} r \syn r : \type{Bit} . R}
						\\ q : \type{Qbit} \vdash q \chk \type{Qbit}
						\\ \cdot; P_0 * (q \mapsto \ket{0}) \prove (q \hookrightarrow -)
					}
					{q : \type{Qbit}; P_0 * (q \mapsto \ket{0}) \vdash r \Leftarrow \kw{measure} q; E_3 \syn r : \type{Bit} . (\exists r : \type{Bit}.R)}
					\\ \cdot \vdash \kw{0} \chk \type{Bit}
				}
				{
					\cdot; P_0 \vdash q \Leftarrow \kw{init 0}; E_{23} \syn r : \type{Bit} . (\exists q : \type{Qbit}.R)
				}
				\\ r : \type{Bit}; i,m ; R \prove \top \multimap \top \wedge Id(r, \kw{0})
			}
			{
				\cdot; \lambda i . \lambda m . i = m \wedge (\top * \lambda m' \top)m \vdash E_1; E_{23} \chk r : \type{Bit} . (\top \multimap \top \wedge Id(r, \kw{0}))
			}
		}
		{\cdot \vdash \kw{do} E \chk \{\top\} r : \type{Bit} \{\top \wedge \kw{Id}(r, \kw{0})\}}
	\end{mathpar}
	\end{minipage}
\end{turn}

@@

\subsection{Hadamard basis states}
\begin{lstlisting}[language=QHaskell]
qplus : {\top} r : Qbit {Id(r, |+\>)}
      = do q <= init 0;
           apply H to q;
           return q

qminus : {\top} r : Qbit {Id(r, |-\>)}
       = do q <= init 1;
            apply H to q;
            return q
\end{lstlisting}

\subsection{Creating entanglement}
\begin{lstlisting}[language=QHaskell]
share : \Pi a : Qbit.
        {a \in {|+\>, |-\>}}   -- a \in {b, c} is short for Id(a,b) \/ Id(a,c)
           b : Qbit
        {Id(a, b) /\ a \in {|0\>, |1\>}}
      = \a.do b <= init 0;
              apply CNOT to [a, b];
              return b
\end{lstlisting}

@@

\subsection{Testing modular Bell pair}
\begin{lstlisting}[language=QHaskell]
testBell : {\top} (a, b) : (Bit o* Bit) {\top /\ Id(a, b)}
         = do (qa, qb) <- bell;
              return (measure qa, measure qb)
\end{lstlisting}

\section{Teleportation}
\label{app:teleport}
\todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{lstlisting}[language=QHaskell]
alice : \Pi a : Qbit. \Pi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bit o* Bit)
        {\top}
      = \a.\e.do apply CNOT to [a, e];
                 apply H to a;
                 return (measure a, measure e)
\end{lstlisting}

\subsection{Bob's circuit}
\begin{lstlisting}[language=QHaskell]
bob : \Pi m1 m2 : Bit. \Pi e : Qbit
      {entangled(e)}
         r : Qbit
      {(Id(r, -)}
      = \m1.\m2.\e.do if m1 then (apply Z to e) else ();
                      if m2 then (apply X to e) else ();
                      return e
\end{lstlisting}
